When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America

### 2020-08-25 code/Submissionv3.Rmd
commit

  _This paper explores variation in the professionalism of opinion pieces authored by retired US military officers and published in major US newspapers in the past 40 years. In doing so, this paper develops a theory that argues that two societal forces, political polarization and military credibility, interact in unique ways that ultimately strain the individual military officer's concepts of loyalty and duty in important ways. Assembling and analyzing more than 330 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1979, the paper argues there has been an increase in published opinion pieces which fail to uphold professional standards as defined by historical military norms. The paper concludes by arguing that today's political polarization reveals a profoundly changing society, and that the norm of a professional, apolitical military will remain under further strain in the current political environment._  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

In late May and early June 2020, racial tension and concerns about police brutality gripped the American public following the death of George Floyd, an unarmed black man, by police in Minneapolis. In several cities and communities across the nation, mostly peaceful protests were followed by riots and looting. On June 1, after giving remarks at the Rose Garden, President Donald Trump walked to St. John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and, several minutes later, returned to the White House.

@@

Additionally, several retired military officers were among the elites who weighed in to give their opinions not only on the Trump administration's handling of specific events near the White House, but also on the broader issues of race in America. But for those paying attention to what these retired officers said, it is also clear that there was substantial variation in the tone and tenor of each officer's remarks. While all of them expressed deep concern about the prospect of active-duty forces confronting peaceful protesters, and thus, violating protections held by the First Amendment of the US Constitution [@brooks_dismay_2020; @dempsey_former_2020], some officers went much further in their criticism, leveling direct insults of the President's fitness to lead [@goldberg_james_2020], his administration's broader policies on race [@allen_moment_2020], and his general leadership abilities [@mullen_i_2020].

This paper examines the way in which retired military officers have attempted to influence public opinion in contemporary America through opinion-editorial publications. It is an interesting question worthy of examination because, as I argue, at least three important factors are interacting in unique ways to shape how and why retired military officers speak out on a variety of issues. These factors are professionalism, political polarization, and military credibility. I argue that the latter two factors, political polarization and institutional credibility, are interacting in unique ways that create tremendous strains on the military's professional nature, and that these factors will continue to do so.

This paper proceeds in four parts. First, the paper briefly reviews the concept of military professionalism in order to orient the reader to the central role this concept has played in the ethos of the American military for the past several decades. Second, the paper builds a theory that demonstrates how the uniqueness of today's political polarization and institutional credibility are interacting to create an environment that threatens professional military behavior in ways that could be highly destructive. Third, the paper explores the data, which consists of more than 330 opinion-editorial pieces published in major US newspapers by retired military officers since the late 1970s. The analysis ultimately shows that although the vast majority of op-eds published by retired military officers maintain professional standards, there is clear evidence that retired military officers are using the pen to both argue about topics that fall outside of the realm of traditional national security topics, and to express clear partisan preferences when making their arguments. Fourth, the paper concludes by briefly exploring the implications of these results on the military as an institution, and specifically, on its ability to remain professional in a domestic setting marked by high political polarization in which the military commands wide public support.

\singlespace
# Military Professionalism as a Foundational Trait
\doublespace
The ideal of the US military acting as an objective, apolitical and non-partisan institution is storied and long. One of the strongest origins of the norm stems from the work of the political scientist, Samuel Huntington, who, in his 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, asserted that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington said generated requirements for the military in that the military could never lose so many aspects of it such that it no longer represented the liberal democracy that it defended [@huntington_soldier_1957, 1-3]. Huntington's point was that America needed a strong military that could win on the battlefield, but cautioned that such a military could not be so foreign in character so as to alienate the character of American liberal society. Huntington argued that the key to resolving the tension generated by these two opposing forces was to adopt a scheme of "objective civilian control" in which political leaders established the ends and goals of policy, while military leaders developed and implemented the military plans to accomplish these goals [@huntington_soldier_1957, 80-85]. To Huntington, largely separate operating spheres between the political leaders of a democracy and the military as an institution would ensure that the military remained "professional," and thus, untainted by partisan interests and concerns. In this way, Huntington posited that the American military would best meet the task of defeating security threats while also remaining apolitical.

While Huntington's theory has enjoyed tremendous staying power (Huntington is still studied and taught in many American military institutions), he is not without critics [see, for example, @feaver_civil-military_1996; @cohen_supreme_2003; @feaver_armed_2005; and @brooks_paradoxes_2020]. It would be too burdensome to unpack all of those here, but one of the earliest and richest critiques of Huntington, and perhaps one of the most relevant given the potential influences of political polarization and institutional credibility, came from S.E. Finer. Authored in 1962, Finer's book, _The Man on Horseback: The Role of the Military in Politics_, argued that Huntington's concept of professionalism simply was not strong enough to keep the military out of politics [@finer_man_1962]. Finer foresaw military officers adopting an ethic of professionalism such that they would actually be more likely, not less likely, to become involved in political issues [@finer_man_1962, 22-23].

Finer was especially concerned that the adoption of professionalism would lead to long-serving military professionals believing that they knew what was best for their country over and above that of temporary elected civilian leaders [@finer_man_1962, 22-23]. Finer warned that "the moment the military draw this distinction between nation and the government in power, they begin to invent their own private notion of the national interest, and from this it is only a skip to the constrained substitution of this view for that of the civilian government…this purported care for the national interest as defined by the military is indeed one of their main reasons for intervening" [@finer_man_1962]. In Finer's view, what was really needed to keep the military out of politics was not only the concept of professionalism, but rather a deep internalization of "the principle of civil supremacy" [@finer_man_1962].^[
Importantly, neither Huntington's concept of professionalism nor Finer's critique of the same are arguing for blind obedience by the military to all civilian orders and directions, a point that is, in the author's opinion, often missed in debates about military professionalism. On the contrary, both Huntington and Finer make room and acknowledge that there are instances when military disobedience is not only required, but justified. However, both agree that in general terms, these situations should be exceedingly rare.]

It is not only in the academic literature that the tradition and worth of an apolitical military has been cultivated. The operational military has also been central in upholding the military's non-partisan traditions, albeit with a far less-than-perfect record. Retired Army General Douglas MacArthur, a hero of World War Two in the Pacific and later, Commander of the occupation of Japan, is well known for his disagreement with President Truman regarding American strategic goals during the Korean War. And yet, the remarks MacArthur made to cadets at West Point in 1962, as an 82 year old retired officer, speak to the primacy of professional military conduct:

@@

Finally, current military doctrine and practices also reinforce the centrality of professional, apolitical, and non-partisan conduct. Prior to the 2016 Presidential Election, Marine Corps General Joseph Dunford, who at the time served as Chairman of the Joint Chiefs of Staff, said that the basis upon which the military is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016]. At nearly the same time, his predecessor, Retired Army General Martin Dempsey, published an opinion-editorial piece in _The Washington Post_ which criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the Republican and Democratic conventions. Dempsey wrote, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016].

Although a professional and non-partisan military is the theoretical goal of the institution, the practical record and results remain far from perfect. From MacArthur's clash with President Truman in the early 1950's, to the so-called "Revolt of the Generals" in 2005-2006, military leaders, even those who are retired, have ventured into political territory on more than one occasion. Nonetheless, academic theorists and military leaders alike have long recognized the importance of professional conduct rooted in non-partisan, apolitical conduct. At its core, such conduct ultimately earns the trust and confidence of the American people.

It must be stressed that the concept of military professionalism does not impose a normative prohibition on the publication of an opinion piece, or the expression of other forms of public dialogue, by a military professional (though there are certainly practical differences between the manner these are employed by those who currently serve and those who are retired). Indeed, Huntington's concept of military professionalism recognizes the unique skills of military officers, some of whom, especially multistar generals and admirals, possess more than 30 and in some cases 40 years of professional experience.

It is reasonable to conclude that a professional opinion piece written by a military officer would take on three very important qualities. First, the military professional would address a relevant military topic. To be clear, multistar generals and admirals often are experienced in the realm of strategy, a word that is often misunderstood and misapplied in discussions of the topic [see for example @strachan_direction_2014]. Without unpacking the concept of strategy further, the relevant point is that military elites, and especially those who have achieved a high rank, have been privy to decisions and experiences occurring near the highest levels of government, and as such, their experiences are not confined to only tactical matters. They are familiar with issues ranging from how wars should be best fought, to future security threats looming on the horizon, to fixing the institutional culture of a military service. Second, from a credibility and expertise perspective, a professional piece would be based on the officer's personal experiences and would be written from an individual's perspective so as not to infer that an argument is being made on behalf of the collective service or element within the military body. Third, an opinion made by a military officer would be free of any hint of partisan sentiment, whether advocacy or denouncement. If one of these three qualities fails to be upheld, there is increasing likelihood that the officer's words can, either intentionally or unintentionally, serve as a weapon, and, like other kinds of weapons, cause great harm. To understand how, we will now turn our attention to understanding the impacts on military professionalism caused by two significant changes that have developed within American Society in the past 50 years. The first is increasing credibility of the military institution, and the second is the introduction of an unparalleled and divisive political polarization. The interaction of these forces is significantly straining the professional conduct of the military institution.

@@

One of the two significant societal changes that has occurred in America during the past fifty years is the increase in credibility of the military institution. Consider that in the late 1970s, as a likely result of public opinion over the Vietnam War, 54% of Americans indicated that they trusted the military "a great deal or quite a lot," whereas in 2019, 73% of Americas expressed the same sentiments [@gallup_polls_confidence_2007]. Compared to other government institutions, the American public trusts the military to a remarkably high degree. According to a recent Pew Research Center poll, 80% of the American public expects the military to act in the best interests of the American public, compared to business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018]. It is accurate to say that America trusts its military more than any other entity.

Military leaders are, rightfully so, cognizant of the benefits of holding the public's trust. In the past several years for example, there has been a concerted effort by the services to identify and punish senior leaders who engage in misconduct, to demonstrate to Congress that senior military leaders do not need assistance to handle sexual assault, and to otherwise prove that it can be trusted to accomplish its missions [@us_army_repairing_2014]. Former Chief of Staff of the Army, Retired General Ray Odierno, told fellow Army generals that "of great importance within the Army Profession is the duty of each of us to be a steward of our profession and to maintain its essential and hard won trust with society [Ray Odierno in @us_army_repairing_2014].

Scholars have also shown that when elites possess a high degree of credibility, their ability to influence and shape public opinion is likewise high. In his influential study, _The Nature and Origins of Mass Opinion_, John Zaller argues that formations in mass public opinion are the result of three interacting components: "variation in the information carried in elite discourse, individual differences in attention to this information, and individual differences in political values and other predispositions" [@zaller_nature_1992]. Zaller argues that throughout modern America, most citizens, as non-elites, must rely heavily on the messages that come from elites regarding various topics and issue areas [@zaller_nature_1992]. Zaller further argues that the messages that elites send is necessarily "stereotyped" in some way: that is, the messages that elites send, in order to be understood by regular citizens who are part of the public, must be "short, simple, and highly thematic" [@zaller_nature_1992]. In developing this argument, Zaller is not claiming that regular citizens are uninformed or careless people who are incapable of processing complex information, but rather acknowledging that most citizens care most about a world that is not "beyond their personal experience" [@zaller_nature_1992]. Thus, his main point is simply that elites play a significant role in forming and swaying public opinion.

@@

The second and decisive societal change that has occurred in the past fifty years is the degree to which and character of the political polarization prevalent in the US political landscape today. Although scholars have offered differing explanations for the causes of this polarization, ranging from concerns over race and religion [@abramowitz_great_2018], to the size and role of government [@sides_two_2015], to fierce policy disagreements over policy issues [@mann_its_2012], virtually all scholars agree that polarization in America has become more pronounced in recent decades [@hare_polarization_2014; @bonica_mapping_2014], and that it is, by many measures, worse than it has been during any period since the end of the American Civil War [@hare_polarization_2014]. The polarized political climate today extends to far more issues than previously polarized periods in American history [@layman_party_2002], is marked by stronger negative feelings towards those who hold opposing views [@abramowitz_great_2018], and, at root, involves fundamental issues of both personal and national identity that are not conducive to any reasonable form of compromise [@abramowitz_great_2018; @mohler_jr_gathering_2020]. Through the lens of these characteristics, one can begin to adequately grasp the challenges of polarization to the professional character of the nation's military.

The first way in which today's highly polarizing political climate is different than that of previous periods in American history is its scope. Fewer and fewer issue areas and policy matters are outside of the "spell" of polarization, owing to a phenomenon known as "conflict extension" [@@layman_party_2002]. In previous eras of polarization, a single issue area served as the "battleground" for partisan conflict. Today, though, this is no longer true. Research has consistently shown that the polarizing politics present in the United States today has caused voters who identify as liberals to be more liberal across multiple issue categories, and that the same holds for conservatives [@abramowitz_great_2018]. This includes the full range of social welfare, cultural, and racial issues, all of which are important in the minds of the electorate and partisans [@layman_party_2006]. As a result, the prospect for cooperation on issues that were not politically polarized remained high, whereas today, the prospect for cooperation in general is lower as a result of the shrinking, non-partisan space [@layman_activists_2010].

The second important characteristic of today's polarizing climate involves the strong, negative feelings that each side espouses towards those who hold differing opinions on a variety of issues. Known as "the rise of negative partisanship" [@abramowitz_great_2018], these strongly negative feelings toward those who hold different political opinions is not a phenomenon that has occurred among political elites only; instead, this trend has been shown to have grown among voters since the late 1970s [@abramowitz_great_2018]. Therefore, in addition to there being a diminished opportunity for compromise owing to the proliferation of issues that are polarized, the feelings produced by today's polarized climate are so strong that the other side believes that other is deeply and morally wrong.

The third characteristic of today's polarizing environment is that polarizing issues do not involve matters of simple policy difference, but rather deep and fundamental beliefs about individual and national identities and purposes [@abramowitz_great_2018; @sides_identity_2018; @mohler_jr_gathering_2020]. The election of Donald Trump and his years in office as President have raised the salience on a host of issues, all of which are essential to various parts of the electorate, including attitudes about race, immigration, and religion  [@sides_identity_2018]. Moreover, the American people are aware of the polarized environment, and in a recent survey, indicated that they are concerned the trend is not improving [@schaeffer_far_2020].

In summary, taken together, these three characteristics of contemporary US political polarization are sobering. As a large body of research indicates, it is right to understand the political climate which American finds itself in today not merely as another polarized period, but rather in a tumultuous era that may well challenge the very definition of what America is and what it stands for [@mann_its_2012; @sides_identity_2018; @abramowitz_great_2018; @mohler_jr_gathering_2020]. How the military may be impacted by the forces of increasing institutional credibility and today's political polarization is the subject of the next section.

@@

Figure 1, below, visually describes a general theory of how institutional credibility and political polarization interact to impact military professionalism. From left to right, two significant societal changes (node 'SC') have occurred since the late 1970s, both of which have been described, and as this theory contends, interact, to impact military professionalism. The first societal change, that of institutional credibility, is indicated by the node 'IC', and the second, that of political polarization, is indicated by the node 'PP'. Each of these, in turn, impact military professionalism, indicated by the node 'MP'. Last, though it is outside of the scope of this paper, I contend that changes in military professionalism as a result of the forces of increasing political polarization and institutional credibility, will impact various military outcomes (node 'MO'). These military outcomes vary, and could include things such as military effectiveness, to include the military's ability to fight and win battles [@talmadge_dictators_2015], as well as strategy formulation, to include the military's ability to work with civilians to produce sound strategic advice [@rapp_civil-military_2015]. Indeed, the full degree to which and in what ways polarization will impact the military are not yet fully known.

@@

```{r, include=FALSE, echo=FALSE}
problem_dag <- dagify(
    MO ~ MP,
    MP ~ PP + IC,
    PP~SC,
    IC~SC,
    coords = tribble(
      ~ name,      ~ x,    ~ y,
      "MO", 1, .5,
      "MP", .7, .5,
      "IC", .3, 1,
      "PP", .3,0,
      "SC", 0, .5
    ))

@@

>>>>

  _This paper explores variation in the professionalism of opinion pieces authored by retired US military officers and published in major US newspapers in the past 40 years. In doing so, this paper develops a theory which argues that increasing political polarization in the United States impacts the military as a profession by straining the individual military officer's concept of loyalty in important ways. Drawing on an original data set of more than 330 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1979, the paper argues there has been an increase in published opinion pieces which fail to uphold professional standards as defined by historical military norms and standards. The paper concludes by arguing that the concept of "politicization of the military" --- something that previous civil-military relations scholars have identified and described --- must be widened to account not only for efforts by politicians to use the military for political advantages, but also for the novel phenomenon of retired military officers using the concept of loyalty to achieve political ends. The author argues that such a development, if uncorrected, will cause significant damage to the military as a profession._  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

In late May and early June 2020, racial tension and concerns about police brutality gripped the American public following the death of George Floyd, an unarmed black man, by police in Minneapolis. In many cities and communities across the nation, largely-peaceful protests were followed by periods of rioting and looting. On June 1, after giving remarks at the Rose Garden, President Donald Trump walked to St. John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and, several minutes later, returned to the White House.

@@

Furthermore, several retired military officers weighed in not only on the Trump administration's handling of specific events near the White House, but also on the broader issues of race in America. For those paying attention to what these retired officers said, it is clear that there was substantial variation in the tone and tenor of each of these officer's remarks. While all of them expressed deep concern about the prospect of active-duty forces confronting peaceful protesters, and thus, violating protections held by the First Amendment of the US Constitution [@brooks_dismay_2020; @dempsey_former_2020], some officers went much further in their criticisms, leveling direct insults of the President's fitness to lead [@goldberg_james_2020], his administration's broader policies on race in America [@allen_moment_2020], and his general leadership abilities [@mullen_i_2020]. Moreover, in the weeks and months since George Floyd's death, retired and currently-serving military officers have further taken up the pen to argue for a variety of topical issues that are in one way or form related to President Donald Trump and his administration's response to a variety of current themes in domestic politics, including race in America [@askew_anti-racist_2020] and the US Postal Service and concerns about the upcoming Presidential Election [@nagl__2020].

This paper examines the way in which retired military officers have attempted to influence public opinion in contemporary America by authoring opinion-editorial publications. It is a question worthy of examination for many reasons, including that at its root, at least three factors are involved, each of which may shape civil-military relations: military professionalism, political polarization, and institutional credibility of the military. The main argument of this paper is that a weakening of the concept of military professionalism can clearly be seen in the way retired military officers have made arguments in the public square over the past four decades, that the main variable that can explain this variation is increasing political polarization within American Society. Further, this paper argues that the military institution is particularly susceptible to increasing political polarization because of the way the concept of loyalty is used to make arguments that contain a clear partisan or political leaning. These developments reveal the extent to which increasing political polarization is synonymous with a fractured domestic public, and highlight the extraordinary challenge that the US military will llikely face to remain a  professional in an enduring polarized political climate.

This paper proceeds in four parts. First, the paper briefly reviews the concept of military professionalism in order to orient the reader to the central role this concept has played in the ethos of the modern American military. Second, the paper builds a theory that demonstrates how certain features of today's polarized political landscape challenge the concept of loyalty for the individual military professional. Third, the paper explores the data, which consists of more than 330 opinion-editorial pieces published in major US newspapers by retired military officers since the late 1970s. This analysis ultimately shows that although the vast majority of op-eds published by retired military officers maintain professional standards, there is clear evidence that retired military officers are using the pen to both argue about topics that fall outside of the realm of traditional national security topics, and to express clear partisan preferences when making their arguments. The analysis also shows that this trend has worsened as the American political climate has become increasingly polarized. Fourth, the paper concludes by briefly exploring the implications of these results on the military as an institution, and specifically, on its ability to remain professional in a highly polarized setting.

\singlespace
# Military Professionalism as a Foundational Trait
\doublespace
The ideal of the US military acting as an objective, apolitical and non-partisan institution is long and storied. As a concept, the idea of a professional military stems from the post-WW2 era, when the US military was locked in the Cold War struggle agains the Soviet Union. Perhaps the strongest and certainly one of the original intellectual founders of a professional military is the political scientist, Samuel Huntington, who, in his 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, asserted that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington said generated requirements for the military in that the military could never lose so many aspects of it such that it no longer represented the liberal democracy that it defended [@huntington_soldier_1957, 1-3]. Huntington's point was that America needed a strong military that could win on the battlefield, but cautioned that such a military could not be so foreign in character so as to alienate the character of American liberal society. Huntington further argued that the key to resolving the tension generated by these two opposing forces was to adopt a scheme of "objective civilian control" in which political leaders established the ends and goals of policy, while military leaders developed and implemented the military plans to accomplish these goals [@huntington_soldier_1957, 80-85]. To Huntington, separate operating spheres between the political leaders of a democracy and the military as an institution would ensure that the military remained "professional," and thus, untainted by partisan interests and concerns. In this way, Huntington posited that the American military would best meet the task of defeating security threats while also remaining apolitical.

While Huntington's theory has enjoyed tremendous staying power (Huntington is still studied and taught in many American military institutions), he is not without critics [see, for example, @feaver_civil-military_1996; @cohen_supreme_2003; @feaver_armed_2005; and @brooks_paradoxes_2020]. It would be too burdensome to unpack all of those here, but one of the earliest and richest critiques of Huntington came from S.E. Finer. Finer's 1962 book, _The Man on Horseback: The Role of the Military in Politics_, argued that Huntington's concept of professionalism simply was not strong enough by itself to keep the military out of politics [@finer_man_1962]. In fact, Finer foresaw military officers adopting an ethic of professionalism such that they would actually be more likely, not less likely, to become involved in political issues [@finer_man_1962, 22-23].

Finer was especially concerned that the adoption of professionalism would lead to long-serving military professionals believing that they knew what was best for their country over and above that of temporary elected civilian leaders [@finer_man_1962, 22-23]. In fact, Finer warned that "the moment the military draw this distinction between nation and the government in power, they begin to invent their own private notion of the national interest, and from this it is only a skip to the constrained substitution of this view for that of the civilian government…this purported care for the national interest as defined by the military is indeed one of their main reasons for intervening" [@finer_man_1962]. Therefore, in Finer's view, what was really needed to keep the military out of politics was not merely adoption of an ethic that relied on professionalism, but rather a deep internalization of "the principle of civil supremacy" [@finer_man_1962].^[
Importantly, neither Huntington's concept of professionalism nor Finer's critique of the same are arguing for blind obedience by the military to all civilian orders and directions, a point that is, in the author's opinion, often missed in debates about military professionalism. On the contrary, both Huntington and Finer make room and acknowledge that there are instances when military disobedience is not only required, but justified. However, both agree that in general terms, these situations should be exceedingly rare.]

It is not only in the academic literature that the tradition and worth of an apolitical military has been cultivated. The operational military has also been central in upholding the military's non-partisan traditions, albeit with a far less-than-perfect record. Retired Army General Douglas MacArthur, a hero of World War Two in the Pacific and later, Commander of the military occupation of Japan, is well known for his disagreement with President Truman regarding American strategic goals during the Korean War. And yet, MacArthur's address to the cadets at West Point in 1962, as an 82 year old retired officer, speaks to the importance that MacArthur felt of the military professional focusing on appropriate military conduct:

@@

Finally, current military doctrine and norms further reinforce the centrality of professional, apolitical, and non-partisan conduct by the military. For example, prior to the 2016 Presidential Election, Marine Corps General Joseph Dunford, who at the time served as Chairman of the Joint Chiefs of Staff, said that the basis upon which the military is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016]. At nearly the same time, his predecessor, Retired Army General Martin Dempsey, published an opinion-editorial piece in _The Washington Post_ which criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the Republican and Democratic conventions. Dempsey wrote, "As generals, they [Flynn and Allen] have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016].

Although a professional and non-partisan military is the theoretical goal of the institution, the practical record and results remain far from perfect. From MacArthur's clash with President Truman in the early 1950's, to the so-called "Revolt of the Generals" in 2005-2006, military leaders, even those who are retired, have ventured into political territory, in one form or another, on more than one occasion. Nonetheless, academic theorists, military tradition, and contemporary military leaders have long argued for the importance of the military remaining grounded in professional conduct that is both non-partisan and apolitical. At its core, it is this type of professional conduct which ultimately postures the military to earn the trust and confidence of the American people.

To be clear, the concept of military professionalism does not impose a normative prohibition on the publication of an opinion piece, or the expression of other forms of public dialogue, by a military professional. In fact, one of the foundations of Huntington's concept of military professionalism is the fact that military leaders possess unique and specialized skills that civilians simply do not have. Moreover, multistar generals and admirals in the contemporary military possess in many cases upwards of 30 and sometimes 40 years of service, and it is entirely reasonable to conclude that these officers in particular have been privy to decisions and experiences occurring at or near the highest levels of government. In other words, their experiences are not confined to only tactical matters. They are familiar with issues ranging from how wars should be best fought, to future security threats looming on the horizon, to fixing the institutional culture of a military service, and more.

It is reasonable, therefore, to conclude that a professional opinion piece written by a military officer would take on three qualities. First, a professionally-written opinion piece would address a relevant military topic consistent with the military expertise of the author. In the US context in particular, where the military conducts a variety of roles and performs a variety of missions, this would include a wide swath of topics, while at the same time, certain topics would be considered out of the realm of relevant military expertise, such as partisan domestic politics. Second, from the perspective of both expertise and institutional credibility, a professional opinion piece would be based on the officer's personal experiences and be written from an individual's perspective so as not to infer that an argument is being made on behalf of the military. Third, a professionally-written opinion piece would be free of of partisan sentiment or expression, whether advocacy or denouncement. If one of these three qualities -- addressing a relevant military topic, written from the perspective of an individual officer's expertise, free of partisan sentiment -- fails to be upheld, there is an increasing likelihood that the officer's words can be conveyed and/or received as unprofessional.

If and when a military officer violates standards of military professionalism when conveying an opinion to the American public, his or her words can cause great harm, either to American society, the military as an instition, or a combination of both. To understand how and why this is the case, we will now turn our attention to understanding the impacts on military professionalism caused by two changes that have developed within American Society in the past 50 years. The first is a gradual increase in the American public's opinion of the military as an institution, and the second, more significant change is the steady increase in political polarization within American society.

@@

The credibility of the American military as an institution has increased during the past fifty years. In the late 1970s, likely as a result of public opinion over the Vietnam War, 54% of Americans indicated that they trusted the military "a great deal or quite a lot," whereas in 2019, 73% of Americas expressed the same sentiments [@gallup_polls_confidence_2007]. Between these time periods, there have been some shorter and localized changes, but by and large, the increase that has occurred has been gradual and steady. Today in fact, relative to other government institutions, the American public trusts the military to a remarkably high degree. For example, according to a recent Pew Research Center poll, 80% of the American public expects the military to act in the best interests of the American public, compared to tho percentage who believe the same about business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018].  
Military leaders are, rightfully so, cognizant of the benefits of holding the American public's trust. Former Chief of Staff of the Army, Retired General Ray Odierno, told fellow Army generals that "of great importance within the Army Profession is the duty of each of us to be a steward of our profession and to maintain its essential and hard won trust with society [Ray Odierno in @us_army_repairing_2014]. For this reason, for example, there has been a concerted effort by the services to identify and punish senior leaders who engage in activities that may damage the "hard won" trust Odierno is referring to, activities such as sexual or financial misconduct [@us_army_repairing_2014].

Scholars have shown that when elites possess a high degree of credibility, their ability to influence and shape public opinion is likewise high. In his influential study, _The Nature and Origins of Mass Opinion_, John Zaller argues that formations in mass public opinion are the result of three interacting components: "variation in the information carried in elite discourse, individual differences in attention to this information, and individual differences in political values and other predispositions" [@zaller_nature_1992]. Zaller argues that throughout modern America, most citizens, as non-elites, must rely heavily on the messages that come from elites regarding various topics and issue areas [@zaller_nature_1992]. Zaller further argues that the messages that elites send is necessarily "stereotyped" in some way: that is, the messages that elites send, in order to be understood by regular citizens who are part of the public, must be "short, simple, and highly thematic" [@zaller_nature_1992]. In developing this argument, Zaller is not claiming that regular citizens are uninformed or careless people who are incapable of processing complex information, but rather acknowledging that most citizens care most about a world that is not "beyond their personal experience" [@zaller_nature_1992]. Thus, his main point is simply that elites play a significant role in forming and swaying public opinion.

@@

The second and more influential societal change that has occurred over the past fifty years in America is the rise of a polarized political landscape. Though this is not the first time in America's history that political polarization has been a dominant feature of domestic politics, multiple scholars have argued that polarization has worsened in recent decades [@hare_polarization_2014; @bonica_mapping_2014]], and some have argued that polarization is at the worse level since the end of the American Civil War [@hare_polarization_2014]. Although scholars have offered differing explanations for the cause of the political polarization evident today, ranging from concerns over race and religion [@abramowitz_great_2018], to the size and role of government [@sides_two_2015], to fierce policy disagreements over policy issues [@mann_its_2012], there is a fair amount of common ground among scholars regarding how polarization has impacted contemporary domestic politics. It is worth unpacking several features of America's polarized landscape, particularly for how these features might alter professional military behavior specifically and civil-military relations more broadly.

The polarized political climate today extends to far more issues than previously polarized periods in American history [@layman_party_2002], is marked by stronger negative feelings towards those who hold opposing views [@abramowitz_great_2018], and, at root, involves fundamental issues of both personal and national identity that are not conducive to any reasonable form of compromise [@abramowitz_great_2018; @mohler_jr_gathering_2020]. Through the lens of these characteristics, one can begin to adequately grasp the challenges of polarization to the professional character of the nation's military.

The first important characteristic of the polarized political climate is its all-encompassing scope. Fewer and fewer issue areas and policy matters are outside of the reaches of polarization, a phenomenon that American politics scholars have dubbed "conflict extension" [@layman_party_2002]. These scholars note that as polarization has increased, more issue areas have become divisive in a partisan sense. The practical effect of "conflict extension" is that the perceived non-partisan or apolitical space has significantly shrunk. In fact, scholars have also shown that consistent with trends of increasing polarization, voters who identify as liberals have become more more liberal across more categories of issues, and that the same trend holds for conservatives [@abramowitz_great_2018]. This includes a range of social welfare, cultural, and racial issues, all of which are important in the minds of the electorate and partisans [@layman_party_2006]. Whereas in previous political eras, the prospect for cooperation on issues that were not politically polarized remained high, the prospect for cooperation across partisan sides is lower as a result of the shrinking, non-partisan space [@layman_activists_2010].

The second important characteristic of today's polarizing climate involves the rise of strong, negative feelings that each side espouses towards those who hold differing opinions on a variety of issues. Known as the rise of "negative partisanship" [@abramowitz_great_2018], citizens holding strongly negative feelings toward those who hold different political opinions is not a phenomenon that has occurred only among political elites; instead, this trend has been shown to have grown among all voters since at least the late 1970s [@abramowitz_great_2018]. The effect of "conflict extension" and "negative partisanship" is a reduction in perceived or actual nonpartisan space as well as a genuine belief that opposing partisans are deeply wrong and flawed.

The third characteristic of today's polarizing environment is that the issues which partisans disagree about do not involve matters of simple policy difference, but rather deep and fundamental beliefs about individual and national identities and purposes [@abramowitz_great_2018; @sides_identity_2018; @mohler_jr_gathering_2020]. Indeed, the election of President Donald Trump and his years in office as President have raised the salience on a host of fundamental issues, many of which are dearly held by various parts of the electorate, including attitudes about race, immigration, and religion  [@sides_identity_2018]. The impact of this particular feature of contemporary political polarization in America is that real moral issues are at stake for partisans on each respective side of the political divide.

Taken together, these three characteristics of contemporary US political polarization offer a sobering assessment of the contemporary political landscape. As a large body of research indicates, it is right to understand that America finds itself today not merely in yet another polarized period, but rather in a tumultuous era that may well challenge the very definition of what America is and what it stands for [@mann_its_2012; @sides_identity_2018; @abramowitz_great_2018; @mohler_jr_gathering_2020]. For anyone - citizens, military officers, or elected officials - to assume that the military will automatically escape these dynamics of polarization simply because of the legacy of the military's professional ethos is to underestimate the strength and all-encompassing nature of the divisions in American society that political polarization underscores. Let us now turn our attention to understanding how the features of today's politically polarized climate may impact the professional nature of the military institution.

@@

Figure 1, below, shows describes a theory for how the variables of political polarization, institutional credibility, military professionalism, and other military outcomes may relate to each other. From left to right, the theory contends that increasing political polarization ultimately reduces military professionalism. At the same time, however, there is an intervening variable, institutional credibility, that may actually increase as a result of increasing political polarization. Finally, on the far right of the diagram, are changes in military outcomes. Though it is outside of the scope of this paper, I contend that eventually, increases in political polarization may at some point cause discernible changes in various military outcomes. These military outcomes may vary, but could include things such as military effectiveness, to include the military's ability to fight and win battles [@talmadge_dictators_2015] and to formulate sound strategic advice [@rapp_civil-military_2015]. Indeed, the full degree to which and in what ways polarization will impact the military are not yet fully known.

@@

```{r, include=FALSE, echo=FALSE}
problem_dag <- dagify(
    MO ~ MP,
    MP ~ PP + IC,
    IC~PP,
    coords = tribble(
      ~ name,      ~ x,    ~ y,
      "MO", 1, .5,
      "MP", .7, .5,
      "IC", .3, .2,
      "PP", .1,.5,
    ))

@@

augment(logmodel1, newdata = predictiondatafut)
